The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-29 manuscript/manuscript.Rnw
combine gene tree figures; fix plot to accept vp

<<plot.defaults,echo=FALSE,results="hide">>=
grad.def <- list(low = "white", high = "green", space = "Lab", na.value = "gray50")

@@

\begin{figure}
<<psg.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PSG1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1)
@ 
\caption{\genetreeexplanation{pregnancy-specific glycoproteins}}
\label{fig:psg_evolution_tree}
\end{figure}

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \bonobo are shown in \cref{tab:top50_bonobo_placenta}
with a list of all genes in Supplemental
\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
placenta differs somewhat from that of \human, as evidenced by the
makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
 in \bonobo is %
\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
%
which is involved in insulin growth factor signaling and angiogenesis
during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
\textit{PRG2} is also expressed at relatively high levels (though
comparatively much lower) in the human placenta
(\Sexpr{gene.expression("homo sapiens","PRG2")};
Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
related to the chorionic somatomammotropins
are also expressed at relatively high levels (GH1 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
GH2
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
%
\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
Placental expression of the genes involved in estrogen and
progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
respectively) show lower expression in \bonobo as compared to \human %
(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
\cref{fig:estrogen_evolution_tree}; %
\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
\cref{fig:progesterone_evolution_tree} and
\cref{tab:top50_human_placenta}) although the expression of
\textit{HSD3B1} is fairly similar to that of \smonkey %
(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
In addition, \textit{CGA} is expressed in the bonobo placenta %
(\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
lower levels than \human as well (\cref{tab:top50_human_placenta}).
%
Two Ensembl-annotated beta subunits of the chorionic gonadotropin
genes are expressed in \bonobo, one related to \textit{CGB} and
another related to \textit{CGB7}; both are not highly expressed
(\cref{fig:cgb_evolution_tree}). Among the other top 50 expressed
genes, there is high expression of keratins (e.g., \textit{KRT19},
\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2}),
insulin growth factor 2 (\textit{IGF2}) and one of its binding
proteins \textit{IGFBP3}, and the hormone adrenomedullin
(\textit{ADM}).

\begin{figure}
<<serpine.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="SERPINE2" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1)
@ 
\caption{\genetreeexplanation{\textit{SERPINE2}}}
\label{fig:serpine_evolution_tree}
\end{figure}

@@

\begin{figure}
<<cathespin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CTSL" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(gene.tree,gene.tree.table,gene.tree.expression,min.fpkm=1,
                 subtree=c("ENSMODG00000002844","ENSMUSG00000055679"))
@ 
\caption{\genetreeexplanation{cathespins related to \textit{CTSL}}}
\label{fig:cathespin_evolution_tree}
\end{figure}

@@

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \carmeli are shown in
\cref{tab:top50_s_carmeli_placenta} with a list of all highly
expressed genes in Supplemental
\cref{tab:all_spalax_carmeli_placenta}. Numerous hemoglobin subunits
ware highly expressed
(\Sexpr{gene.expression("spalax_carmeli","Hbb-bh1",complete=TRUE)},
\Sexpr{gene.expression("spalax_carmeli","Hbb-a1",complete=TRUE)}, and
\Sexpr{gene.expression("spalax_carmeli","Hbb-bs",complete=TRUE)}), as
well as the relaxin
\Sexpr{gene.expression("spalax_carmeli","Rln1",complete=TRUE)},
multiple cathespins
(\Sexpr{gene.expression("spalax_carmeli","Ctsm",complete=TRUE)},
\Sexpr{gene.expression("spalax_carmeli","BC051665",complete=TRUE)},
and \Sexpr{gene.expression("spalax_carmeli","Ctsl",complete=TRUE)},
\cref{fig:cathespin_evolution_tree}), and many prolactin-related genes
(\Sexpr{gene.expression("spalax_carmeli","Prl4a1",complete=TRUE)},
\Sexpr{gene.expression("spalax_carmeli","Prl6a1",complete=TRUE)},
\Sexpr{gene.expression("spalax_carmeli","Prl3b1",complete=TRUE)}, and
\Sexpr{gene.expression("spalax_carmeli","Prl7c1",complete=TRUE)},
\cref{fig:prolactin_evolution_tree}) which shows a similar pattern of
expansion seen in \mouse.

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of the \cow are shown in \cref{tab:top50_cow_placenta} with a
list of all genes in Supplemental \cref{tab:all_bos_taurus_placenta}.
The gene with the highest expression was \textit{TMSB10} %
(\Sexpr{gene.expression("bos taurus","TMSB10")}), and
\textit{TMSB4} was also highly expressed %
(\Sexpr{gene.expression("bos taurus","TMSB4")}). As in
\dog{}, members of annexin complexes were also highly expressed
(\Sexpr{gene.expression("bos taurus","S100G",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ANXA2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","S100A2",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","S100A11",complete=TRUE)}), as well as %
at least one of a set of four miRNA with 100\% identity
(\Sexpr{gene.expression("bos taurus","ENSBTAG00000047046",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000047154",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000045710",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000046351",complete=TRUE)}). 
%
As these are novel miRNA, there is a possibility that the assignment
of the short reads to these four miRNA is due to miss-annotation of
the \cow genome.
% 
Several members of the ruminant-specific pregnancy-associated
glycoprotein family show especially high expression in \cow placenta
(\cref{tab:top50_cow_placenta}, \cref{fig:pag2_evolution_tree}, and
Supplemental \cref{tab:all_bos_taurus_placenta}) including
%
\Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000022759",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000013214",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000019226",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000024611",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG12",complete=TRUE)}. %
%
28 members of this family have $FPKM \ge 1$, and this family of
genes has greatly expanded in ruminants (\cow and \sheep{}) although
the expression levels of this family are higher in \cow than in \sheep
\cref{fig:pag2_evolution_tree}. Other genes with high expression
include genes involved with regulation of growth (\textit{IGF2},
\textit{PHLDA2}), imprinted genes (\textit{DLK1}, \textit{PHLDA2}),
and \textit{CSH1}, a gene that is a member of a ruminant specific
expansion of placental prolactins and not directly related to the gene
of the same name in hominids \ref{fig:prolactin_evolution_tree}. Other
ruminant-specific placental prolactins also show high expression in
the cow placenta including %
\Sexpr{gene.expression("bos taurus","PRP1",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP5",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP6",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP8",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","PRP7",complete=TRUE)}, %
with a total of 14 with $\mathrm{FPKM} \ge 1$.
%
\textit{SSLP1} (\textit{SOLD1} in \sheep{}), a gene expressed in the
cotyledonary villi of the ruminiant placenta and involved in cell
proliferation
\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, also
shows evidence of high expression %
(\Sexpr{gene.expression("bos taurus","SSLP1")}). %
There is some placental expression of the estrogen synthase gene
(\textit{CYP19A1}) in cows %
(\Sexpr{gene.expression("bos taurus","CYP19A1")}), and
comparatively greater expression of \textit{HSD3B} %
(\Sexpr{gene.expression("bos taurus","HSD3B")}).

\begin{figure}
<<pag2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PAG2" & species=="bos taurus",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
                 gene.tree.table,gene.tree.expression,min.fpkm=1,
                 subtree=c("ENSSSCG00000006792",
                           "ENSBTAG00000046085")
                 )
@ 
\caption{\genetreeexplanation{\cow \textit{PAG2}}}
\label{fig:pag2_evolution_tree}
\end{figure}

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \elephant are shown in \cref{tab:top50_elephant_placenta}
with a list of all genes in
\cref{tab:all_loxodonta_africana_placenta}. The gene with the highest
expression was the natriuretic peptide precursor B
(\Sexpr{gene.expression("loxodonta_africana","NPPB",complete=TRUE)}).
Two thymosins,
\Sexpr{gene.expression("loxodonta_africana","TMSB10",complete=TRUE)}
and
\Sexpr{gene.expression("loxodonta_africana","TMSB4Y",complete=TRUE)}
were also also was highly expressed. The miRNA mir-675
(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000027406",complete=TRUE)}).
which is imbedded in the first exon of the maternally imprinted long
non-coding RNA gene \textit{H19}
\cite{Keniry.ea2012:H19lincRNAisdevelopmental}, a gene which is
currently un-annotated in \elephant{}. \DLA{Probably consider fixing
  the unannotation of H19; it seems pretty obvious to me, though it
  would require fixing the GTF file for elephant with the right
  \href{http://useast.ensembl.org/Loxodonta_africana/Transcript/Summary?db=core;pt=GENSCAN00000034833;r=scaffold_145:855987-899506}{genscan
    prediction} for this region} Five uncharacterized genes are
expressed to a high degree in the \elephant placenta
(\cref{tab:top50_elephant_placenta}), three of which
(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000028493",complete=TRUE)},
\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032011",complete=TRUE)},
\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000031933",complete=TRUE)})
are members of an \elephant{}-specific expansion of 11 genes closely
related to the thymus-specific serine protease gene \textit{PRSS16}
\cref{fig:prss16_evolution_tree} The other 8 \textit{PRSS16}-like
genes are also expressed to a lesser degree
(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000026505",complete=TRUE)}).
Two other novel genes with high expression are related to the gene
\textit{SSLP1}
(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032659",complete=TRUE)}
and
\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000029551",complete=TRUE)},
\cref{fig:sslp1_evolution_tree}), which is also expressed in \cow and
\dog placentas (\cref{tab:top50_dog_placenta,tab:top50_cow_placenta}).
Placental expression of the hormones relaxin (\textit{RLN}, %
\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000036090")})
and prolactin (\textit{PRL}) are exceptionally high in the elephant
when compared to other mammals
(\cref{fig:rln2_evolution_tree,fig:prolactin_evolution_tree}). There
is little to no placental expression of the estrogen synthase gene,
\textit{CYP19A1}
(\Sexpr{gene.expression("loxodonta_africana","CYP19A1")}) and the
enzyme that produces progesterone HSD3B1
(\Sexpr{gene.expression("loxodonta_africana","HSD3B")}). Other genes
with high expression in the elephant placenta include genes involved
in oxidative stress
(\Sexpr{gene.expression("loxodonta_africana","TXN",complete=TRUE)}),
growth
(\Sexpr{gene.expression("loxodonta_africana","IGF2",complete=TRUE)})
and
\Sexpr{gene.expression("loxodonta_africana","PAEP",complete=TRUE)}, a
gene usually expressed in the endometrium of the uterus in humans
\cite{Rutanen.ea1987:Progesterone-associatedproteinsPP12and}.

\begin{figure}
<<prss16.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{\genetreeexplanation{\textit{PRSS16}}}
\label{fig:prss16_evolution_tree}
\end{figure}

>>>>

<<plot.defaults,echo=FALSE,results="hide">>=
grad.def <- list(low = "#132B43", high = "#56B1F7", space = "Lab", na.value = "gray50")

@@

\begin{figure}
<<psg.ctsl.evolution.tree,echo=FALSE,fig.width=16,fig.height=16,warning=FALSE,message=FALSE>>=
plot.label.size <- 3

pushViewport(viewport(layout=grid.layout(nrow=2,ncol=2)))
tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PSG1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1,
                 vp=viewport(layout.pos.col=1,layout.pos.row=1),
                 fontsize=plot.label.size
                 )

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CTSL" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(gene.tree,gene.tree.table,gene.tree.expression,min.fpkm=1,
                 subtree=c("ENSMODG00000002844","ENSMUSG00000055679"),
                 vp=viewport(layout.pos.col=2,layout.pos.row=1),
                 fontsize=plot.label.size
                 )

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 vp=viewport(layout.pos.col=1,layout.pos.row=2),
                 fontsize=plot.label.size
                 )

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PAG2" & species=="bos taurus",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
                 gene.tree.table,gene.tree.expression,min.fpkm=1,
                 subtree=c("ENSSSCG00000006792",
                           "ENSBTAG00000046085"),
                 vp=viewport(layout.pos.col=2,layout.pos.row=2),
                 fontsize=plot.label.size
                 )

popViewport(1)
grid.text(c("A","B","C","D"),
          x=unit(c(0,0.5,0,0.5),"npc"),y=unit(c(1,1,0.5,0.5),"npc"),
          just=c("left","top"),
          gp=gpar(fontsize="48")
          )
@ 
\caption{\genetreeexplanation{(A) \textit{PSG1}, (B) \textit{CTSL}, (C) \textit{PRSS16}, and (D) \textit{PAG2}}}
\label{fig:pag2_prss16_ctsl_psg_evolution_tree}
\end{figure}

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \bonobo are shown in \cref{tab:top50_bonobo_placenta}
with a list of all genes in Supplemental
\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
placenta differs somewhat from that of \human, as evidenced by the
makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
 in \bonobo is %
\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
%
which is involved in insulin growth factor signaling and angiogenesis
during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
\textit{PRG2} is also expressed at relatively high levels (though
comparatively much lower) in the human placenta
(\Sexpr{gene.expression("homo sapiens","PRG2")};
Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
related to the chorionic somatomammotropins
are also expressed at relatively high levels (GH1 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
GH2
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
%
\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
Placental expression of the genes involved in estrogen and
progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
respectively) show lower expression in \bonobo as compared to \human %
(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
\cref{fig:estrogen_evolution_tree}; %
\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
\cref{fig:progesterone_evolution_tree} and
\cref{tab:top50_human_placenta}) although the expression of
\textit{HSD3B1} is fairly similar to that of \smonkey %
(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
In addition, \textit{CGA} is expressed in the bonobo placenta %
(\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
lower levels than \human as well (\cref{tab:top50_human_placenta}).
%
Two Ensembl-annotated beta subunits of the chorionic gonadotropin
genes are expressed in \bonobo, one related to \textit{CGB} and
another related to \textit{CGB7}; both are not highly expressed
(\cref{fig:cgb_evolution_tree}). Among the other top 50 expressed
genes, there is high expression of keratins (e.g., \textit{KRT19},
\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2},
Supplemental \cref{fig:serpine_evolution_tree}), insulin growth factor
2 (\textit{IGF2}) and one of its binding proteins \textit{IGFBP3}, and
the hormone adrenomedullin (\textit{ADM}).

@@

@@

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \carmeli are shown in
\cref{tab:top50_s_carmeli_placenta} with a list of all highly
expressed genes in Supplemental
\cref{tab:all_spalax_carmeli_placenta}. Numerous hemoglobin subunits
ware highly expressed
(\Sexpr{gene.expression("spalax_carmeli","Hbb-bh1",complete=TRUE)},
\Sexpr{gene.expression("spalax_carmeli","Hbb-a1",complete=TRUE)}, and
\Sexpr{gene.expression("spalax_carmeli","Hbb-bs",complete=TRUE)}), as
well as the relaxin
\Sexpr{gene.expression("spalax_carmeli","Rln1",complete=TRUE)},
multiple cathespins
(\Sexpr{gene.expression("spalax_carmeli","Ctsm",complete=TRUE)},
\Sexpr{gene.expression("spalax_carmeli","BC051665",complete=TRUE)},
and \Sexpr{gene.expression("spalax_carmeli","Ctsl",complete=TRUE)},
\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~B), and many prolactin-related genes
(\Sexpr{gene.expression("spalax_carmeli","Prl4a1",complete=TRUE)},
\Sexpr{gene.expression("spalax_carmeli","Prl6a1",complete=TRUE)},
\Sexpr{gene.expression("spalax_carmeli","Prl3b1",complete=TRUE)}, and
\Sexpr{gene.expression("spalax_carmeli","Prl7c1",complete=TRUE)},
\cref{fig:prolactin_evolution_tree}) which shows a similar pattern of
expansion seen in \mouse.

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of the \cow are shown in \cref{tab:top50_cow_placenta} with a
list of all genes in Supplemental \cref{tab:all_bos_taurus_placenta}.
The gene with the highest expression was \textit{TMSB10} %
(\Sexpr{gene.expression("bos taurus","TMSB10")}), and
\textit{TMSB4} was also highly expressed %
(\Sexpr{gene.expression("bos taurus","TMSB4")}). As in
\dog{}, members of annexin complexes were also highly expressed
(\Sexpr{gene.expression("bos taurus","S100G",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ANXA2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","S100A2",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","S100A11",complete=TRUE)}), as well as %
at least one of a set of four miRNA with 100\% identity
(\Sexpr{gene.expression("bos taurus","ENSBTAG00000047046",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000047154",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000045710",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000046351",complete=TRUE)}). 
%
As these are novel miRNA, there is a possibility that the assignment
of the short reads to these four miRNA is due to miss-annotation of
the \cow genome.
% 
Several members of the ruminant-specific pregnancy-associated
glycoprotein family show especially high expression in \cow placenta
(\cref{tab:top50_cow_placenta}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D, and
Supplemental \cref{tab:all_bos_taurus_placenta}) including
%
\Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000022759",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000013214",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000019226",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000024611",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG12",complete=TRUE)}. %
%
28 members of this family have $FPKM \ge 1$, and this family of
genes has greatly expanded in ruminants (\cow and \sheep{}) although
the expression levels of this family are higher in \cow than in \sheep
\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D. Other genes with high expression
include genes involved with regulation of growth (\textit{IGF2},
\textit{PHLDA2}), imprinted genes (\textit{DLK1}, \textit{PHLDA2}),
and \textit{CSH1}, a gene that is a member of a ruminant specific
expansion of placental prolactins and not directly related to the gene
of the same name in hominids \ref{fig:prolactin_evolution_tree}. Other
ruminant-specific placental prolactins also show high expression in
the cow placenta including %
\Sexpr{gene.expression("bos taurus","PRP1",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP5",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP6",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP8",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","PRP7",complete=TRUE)}, %
with a total of 14 with $\mathrm{FPKM} \ge 1$.
%
\textit{SSLP1} (\textit{SOLD1} in \sheep{}), a gene expressed in the
cotyledonary villi of the ruminiant placenta and involved in cell
proliferation
\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, also
shows evidence of high expression %
(\Sexpr{gene.expression("bos taurus","SSLP1")}). %
There is some placental expression of the estrogen synthase gene
(\textit{CYP19A1}) in cows %
(\Sexpr{gene.expression("bos taurus","CYP19A1")}), and
comparatively greater expression of \textit{HSD3B} %
(\Sexpr{gene.expression("bos taurus","HSD3B")}).

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \elephant are shown in \cref{tab:top50_elephant_placenta}
with a list of all genes in
\cref{tab:all_loxodonta_africana_placenta}. The gene with the highest
expression was the natriuretic peptide precursor B
(\Sexpr{gene.expression("loxodonta_africana","NPPB",complete=TRUE)}).
Two thymosins,
\Sexpr{gene.expression("loxodonta_africana","TMSB10",complete=TRUE)}
and
\Sexpr{gene.expression("loxodonta_africana","TMSB4Y",complete=TRUE)}
were also also was highly expressed. The miRNA mir-675
(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000027406",complete=TRUE)}).
which is imbedded in the first exon of the maternally imprinted long
non-coding RNA gene \textit{H19}
\cite{Keniry.ea2012:H19lincRNAisdevelopmental}, a gene which is
currently un-annotated in \elephant{}. \DLA{Probably consider fixing
  the unannotation of H19; it seems pretty obvious to me, though it
  would require fixing the GTF file for elephant with the right
  \href{http://useast.ensembl.org/Loxodonta_africana/Transcript/Summary?db=core;pt=GENSCAN00000034833;r=scaffold_145:855987-899506}{genscan
    prediction} for this region} Five uncharacterized genes are
expressed to a high degree in the \elephant placenta
(\cref{tab:top50_elephant_placenta}), three of which
(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000028493",complete=TRUE)},
\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032011",complete=TRUE)},
\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000031933",complete=TRUE)})
are members of an \elephant{}-specific expansion of 11 genes closely
related to the thymus-specific serine protease gene \textit{PRSS16}
\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~C The other 8 \textit{PRSS16}-like
genes are also expressed to a lesser degree
(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000026505",complete=TRUE)}).
Two other novel genes with high expression are related to the gene
\textit{SSLP1}
(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032659",complete=TRUE)}
and
\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000029551",complete=TRUE)},
\cref{fig:sslp1_evolution_tree}), which is also expressed in \cow and
\dog placentas (\cref{tab:top50_dog_placenta,tab:top50_cow_placenta}).
Placental expression of the hormones relaxin (\textit{RLN}, %
\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000036090")})
and prolactin (\textit{PRL}) are exceptionally high in the elephant
when compared to other mammals
(\cref{fig:rln2_evolution_tree,fig:prolactin_evolution_tree}). There
is little to no placental expression of the estrogen synthase gene,
\textit{CYP19A1}
(\Sexpr{gene.expression("loxodonta_africana","CYP19A1")}) and the
enzyme that produces progesterone HSD3B1
(\Sexpr{gene.expression("loxodonta_africana","HSD3B")}). Other genes
with high expression in the elephant placenta include genes involved
in oxidative stress
(\Sexpr{gene.expression("loxodonta_africana","TXN",complete=TRUE)}),
growth
(\Sexpr{gene.expression("loxodonta_africana","IGF2",complete=TRUE)})
and
\Sexpr{gene.expression("loxodonta_africana","PAEP",complete=TRUE)}, a
gene usually expressed in the endometrium of the uterus in humans
\cite{Rutanen.ea1987:Progesterone-associatedproteinsPP12and}.
